Can We Formalise Type Theory Intrinsically without Any Compromise?

### 2025-12-03 tex/CPP26/final/main.tex
submitted camera-ready version

\documentclass[sigplan,screen,natbib,pbalance]{acmart}
\settopmatter{printfolios=true,printccs=false,printacmref=false}
\setcopyright{acmlicensed}
\copyrightyear{2025}
\acmYear{2025}
\acmDOI{XXXXXXX.XXXXXXX}
\acmConference[CPP '26]{Certified Programs and Proofs}{January 12--13, 2026}{Rennes, France}
%%
%%  Uncomment \acmBooktitle if the title of the proceedings is different
%%  from ``Proceedings of ...''!
%%
%%\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
%%  June 03--05, 2018, Woodstock, NY}
\acmISBN{978-1-4503-XXXX-X/2018/06}

%%% Packages %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[utf8]{inputenc}
\usepackage[UKenglish]{babel}
\usepackage{newunicodechar}
\usepackage{xspace}
\usepackage{xcolor}
%\usepackage{natbib}
\setcitestyle{numbers,sort&compress}
\usepackage{mathtools}
\let\Bbbk\relax
\usepackage{amsmath,amssymb,mathbbol}
\usepackage{bbold}
\usepackage[inline]{enumitem}

%% Remove the following if there are no todo items.
\setlength {\marginparwidth }{2cm}
\usepackage[obeyFinal,textsize=footnotesize]{todonotes}
\usepackage[capitalise]{cleveref}
\newcommand{\LT}[2][]{\todo[inline,author={L-T},caption={},color={pink},#1]{#2}}
\newcommand{\FNF}[2][]{\todo[inline,author={Fred},caption={},#1]{#2}}

\usepackage{microtype}
\usepackage{newunicodechar}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\RequirePackage[T1]{fontenc}
\RequirePackage[utf8]{inputenc}
\RequirePackage{amsfonts}
\RequirePackage{newunicodechar}

@@

\DeclareUnicodeCharacter{2115}{\ensuremath{\mathbb{N}}}
\DeclareUnicodeCharacter{2236}{:}
\DeclareUnicodeCharacter{2237}{\ensuremath{\mathrel{::}}}
\DeclareUnicodeCharacter{03A3}{\ensuremath{\Sigma}}
\DeclareUnicodeCharacter{03BB}{\ensuremath{\lambda}}
\DeclareUnicodeCharacter{039B}{\ensuremath{\Lambda}}
\DeclareUnicodeCharacter{03C0}{\ensuremath{\pi}}
\DeclareUnicodeCharacter{03C3}{\ensuremath{\sigma}}
\DeclareUnicodeCharacter{03C9}{\ensuremath{\omega}}
\newunicodechar{β}{\ensuremath{\beta}}
\newunicodechar{η}{\ensuremath{\eta}}
\newunicodechar{Γ}{\ensuremath{\Gamma}}
\newunicodechar{Δ}{\ensuremath{\Delta}}
\newunicodechar{Θ}{\ensuremath{\Theta}}
\newunicodechar{Ξ}{\ensuremath{\Xi}}
\newunicodechar{Π}{\ensuremath{\Pi}}
\newunicodechar{τ}{\ensuremath{\tau}}
\newunicodechar{γ}{\ensuremath{\gamma}}
\newunicodechar{δ}{\ensuremath{\delta}}
\newunicodechar{∅}{\ensuremath{\emptyset}}
\newunicodechar{≡}{\ensuremath{\equiv}}
\newunicodechar{∘}{\ensuremath{\circ}}
\newunicodechar{𝔹}{\ensuremath{\mathbb{B}}}
\newunicodechar{𝕓}{\ensuremath{\mathbb{b}}}
\newunicodechar{∎}{\ensuremath{\blacksquare}}
\newunicodechar{∈}{\ensuremath{\in}}
\newunicodechar{⋆}{\ensuremath{\star}}
\newunicodechar{⌜}{\ensuremath{\ulcorner}}
\newunicodechar{⌝}{\ensuremath{\urcorner}}
\newunicodechar{ʸ}{\ensuremath{{}^y}}
\newunicodechar{ⁿ}{\ensuremath{{}^n}}
\newunicodechar{ᴾ}{\ensuremath{{}^P}}

@@

\DeclareUnicodeCharacter{1D57}{\ensuremath{^t}}
\newunicodechar{∙}{\ensuremath{\cdot}}

@@

\renewcommand\Varid[1]{\mathord{\textsf{#1}}}
%\let\Conid\Varid
%\newcommand\Keyword[1]{\textsf{\textbf{#1}}}

@@

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Can We Formalise Type Theory Intrinsically without Any Compromise?}
\subtitle{A Case Study in \CA}
%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{Liang-Ting Chen}
\affiliation{%
  \institution{Academia Sinica}
  \city{Taipei}
  \country{Taiwan}}
\email{liangtingchen@as.edu.tw}

\author{Fredrik Nordvall Forsberg}
\affiliation{%
  \institution{University of Strathclyde}
  \city{Glasgow}
  \country{United Kingdom}}
\email{fredrik.nordvall-forsberg@strath.ac.uk}

\author{Tzu-Chun Tsai}
\affiliation{%
  \institution{Academia Sinica}
  \city{Taipei}
  \country{Taiwan}}
%\affiliation{%
%  \institution{University of Amsterdam}
%  \city{Amsterdam}
%  \country{Netherlands}}

%%
%% By default, the full list of authors will be used in the page
%% headers. Often, this list is too long, and will overlap
%% other information printed in the page headers. This command allows
%% the author to define a more concise list
%% of authors' names for this purpose.
%\renewcommand{\shortauthors}{Trovato et al.}

@@

\keywords{Proof Assistants, Formalisation, Cubical Agda, Quotient Inductive-Inductive-Recursive Type}
%% A "teaser" image appears between the author and affiliation
%% information and the body of the document, and typically spans the
%% page.

%\received{20 February 2007}
%\received[revised]{12 March 2009}
%\received[accepted]{5 June 2009}

@@

%\LT[inline]{Clearly articulate what has been formalised and what has not, when appropriate? Ideally finish formalisation, otherwise scale back and discuss what has been done. (R1, R3)}
%\LT[inline]{Alternative formulation? E.g., \url{https://web.archive.org/web/20241004151846/https://lists.chalmers.se/pipermail/agda/2019/011176.html} (R2)}

%\listoftodos

@@

\section{Setting and metatheory}

%\LT{Should we discharge this assumption in our formalisation, eg set-truncate where needed instead? (R1, R2)}
Our formalisation is carried out in \CA without the use of the \ensuremath{\Conid{Glue}} type and, in particular, the univalence principle, checked using the option \text{\ttfamily \char45{}\char45{}cubical\char61{}no\char45{}glue} available in the forthcoming \Agda 2.9.0. It also typechecks with \Agda version 2.8.0.
%
We explicitly set-truncate the types we define.
Therefore, the term \emph{set} is used interchangeably with type.
%
For the sake of simplicity we occasionally postulate uniqueness of identity proofs (UIP) locally, which is of course inconsistent with univalence but in principle compatible with cubical type theory.
%
We believe that the cubical type theory XTT~\cite{Sterling2022}, which enjoys definitional UIP without univalence, justifies this local assumption.
%

\CA implements cubical type theory, and one of the important concepts therein is the interval type \ensuremath{\Conid{I}} with two distinguished endpoints \ensuremath{\Varid{i0}} and \ensuremath{\Varid{i1}}.
%
Propositional equality \ensuremath{\Varid{x}\;\mathop\equiv\;\Varid{y}} in a type \ensuremath{\Conid{A}} is given by \emph{paths} in \ensuremath{\Conid{A}}, i.e., by a functions \ensuremath{\Conid{I}\;\Varid{→}\;\Conid{A}}.
More generally, dependent paths \ensuremath{\Varid{p}\;\mathbin{:}\;\Conid{PathP}\;\Conid{P}\;\Varid{a}\;\Varid{b}} are dependent functions \ensuremath{\Varid{p}\;\mathbin{:}\;(\Varid{i}\;\mathbin{:}\;\Conid{I})\;\Varid{→}\;\Conid{P}\;\Varid{i}} sending \ensuremath{\Varid{i0}} to \ensuremath{\Varid{a}\;\mathbin{:}\;\Conid{P}\;\Varid{i0}} and \ensuremath{\Varid{i1}} to \ensuremath{\Varid{b}\;\mathbin{:}\;\Conid{P}\;\Varid{i1}}.
Note that \ensuremath{\Conid{P}\;\mathbin{:}\;\Conid{I}\;\Varid{→}\;\Conid{Type}} itself is a path in the universe \ensuremath{\Conid{Type}}, hence a witness that \ensuremath{\Conid{P}\;\Varid{i0}\;\mathop\equiv\;\Conid{P}\;\Varid{i1}}, which the dependent path is \emph{over}, so paths are special cases of \ensuremath{\Conid{PathP}} with \ensuremath{\Conid{P}} being a constant path.
%
The constant path \ensuremath{\Varid{refl}} defined as \ensuremath{\Varid{λ}\;\anonymous \;\Varid{→}\;\Varid{a}} witnesses that equality is reflexive \ensuremath{\Varid{a}\;\mathop\equiv\;\Varid{a}}, and paths can be lifted to type families in the sense that there is a transport operation \ensuremath{\Varid{subst}\;\mathbin{:}\;(\Conid{P}\;\mathbin{:}\;\Conid{A}\;\Varid{→}\;\Conid{Type})\;\Varid{→}\;\Varid{x}\;\mathop\equiv\;\Varid{y}\;\Varid{→}\;\Conid{P}\;\Varid{x}\;\Varid{→}\;\Conid{P}\;\Varid{y}}.
%
See the literature on cubical type theory for details~\cite{Vezzosi2021}.
%

@@

\section{Type theory as quotient inductive types} \label{sec:tt}

%\FNF{Discuss ``another approach is staying completely indexed and always using PathP instead of transports'' here, or in Discussion (R2)}

@@

%This section's aim is to exhibit that Altenkirch and Kaposi's representation, which contains the transport hells and violates the syntactic restriction imposed by \CA, can be transformed to a representation based on Awodey's natural model, which is free from transports and accepted by \CA.
%Then, we will show how other type formers can be represented in this way.
%In the reminder of this section, we will give its elimination principle and explain how these definitions are formalised in \CA.

\subsection{Type theory as the initial CwF model} \label{sec:tt:cwf}

@@

The four types of judgements in type theory are represented inductive-inductively %and indexed by their context and by their types for terms

@@

When formulating the corresponding rule for the interaction between \ensuremath{\Varid{\char95 ∘\char95 }} and \ensuremath{\Varid{\char95 ,\char95 }}, we encounter a type mismatch that needs to be resolved by a transport (highlighted in \highlight{addition}{\text{green}}): %, leading to the transport hell when reasoning with this equality:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{,∘}\;{}\<[5]%
\>[5]{}\mathbin{:}\;(\Varid{σ}\;\Varid{,}\;\Varid{t})\;\Varid{∘}\;\Varid{τ}\;\mathop\equiv\;(\Varid{σ}\;\Varid{∘}\;\Varid{τ}\;\Varid{,}\;\highlight{addition}{\Varid{subst}\;(\Conid{Tm}\;\Conid{Γ})\;[\circ]_{\text{T}}}\;(\Varid{t}\;[\kern-1.5pt\;\Varid{τ}\;\kern-1.5pt]_{\text{t}})){}\<[E]%
\ColumnHook
\end{hscode}\resethooks
The reason is that the type of \ensuremath{\Varid{t}\;[\kern-1.5pt\;\Varid{τ}\;\kern-1.5pt]_{\text{t}}} is \ensuremath{\Conid{A}\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{T}}\;[\kern-1.5pt\;\Varid{τ}\;\kern-1.5pt]_{\text{T}}} rather than the required \ensuremath{\Conid{A}\;[\kern-1.5pt\;\Varid{σ}\;\Varid{∘}\;\Varid{τ}\;\kern-1.5pt]_{\text{T}}}.
However, since \ensuremath{\Conid{Tm}} is an argument to \ensuremath{\Varid{subst}}, the use of transport violates a syntactic restriction of \Agda, namely its strict positivity check.
In theory, transports are allowed in QIITs~\cite{Kaposi2019}, but it is not clear to us how this syntactic restriction should be relaxed for higher inductive types supported by \CA to take into account other cubical primitives (such as \ensuremath{\Varid{hcomp}}).

@@

\subsection{The `Ford transformation' and index elimination} \label{sec:tt:terms-without-indices}

>>>>

\documentclass[sigplan,screen,natbib,balance]{acmart}

%%%%% Article setup %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% The following is specific to CPP '26 and the paper
%%% 'Can We Formalise Type Theory Intrinsically without Any Compromise? A Case Study in Cubical Agda'
%%% by Liang-Ting Chen, Fredrik Nordvall Forsberg, and Tzu-Chun Tsai.
%%%
\setcopyright{cc}
\setcctype{by}
\acmDOI{10.1145/3779031.3779090}
\acmYear{2026}
\copyrightyear{2026}
\acmISBN{979-8-4007-2341-4/2026/01}
\acmConference[CPP '26]{Proceedings of the 15th ACM SIGPLAN International Conference on Certified Programs and Proofs}{January 12--13, 2026}{Rennes, France}
\acmBooktitle{Proceedings of the 15th ACM SIGPLAN International Conference on Certified Programs and Proofs (CPP '26), January 12--13, 2026, Rennes, France}
\received{2025-09-13}
\received[accepted]{2025-11-13}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% Packages %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[utf8]{inputenc}
\usepackage[UKenglish]{babel}
%\usepackage{newunicodechar}
\usepackage{xspace}
\usepackage{xcolor}
%\usepackage{natbib}
\setcitestyle{numbers,sort&compress}
\usepackage{mathtools}
\let\Bbbk\relax
\usepackage{amsmath,amssymb}
%\usepackage{mathbbol} % not supported by ACM
\usepackage{bbold}
\usepackage[inline]{enumitem}
\usepackage[capitalise]{cleveref}
\usepackage{microtype}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\RequirePackage[T1]{fontenc}
\RequirePackage[utf8]{inputenc}
\RequirePackage{amsfonts}
%\RequirePackage{newunicodechar}

@@

\DeclareUnicodeCharacter{2115}{\ensuremath{\mathbb{N}}}
\DeclareUnicodeCharacter{2236}{:}
\DeclareUnicodeCharacter{2237}{\ensuremath{\mathrel{::}}}
\DeclareUnicodeCharacter{03A3}{\ensuremath{\Sigma}}
\DeclareUnicodeCharacter{03BB}{\ensuremath{\lambda}}
\DeclareUnicodeCharacter{039B}{\ensuremath{\Lambda}}
\DeclareUnicodeCharacter{03C0}{\ensuremath{\pi}}
\DeclareUnicodeCharacter{03C3}{\ensuremath{\sigma}}
\DeclareUnicodeCharacter{03C9}{\ensuremath{\omega}}

\DeclareUnicodeCharacter{03B2}{\ensuremath{\beta}}
\DeclareUnicodeCharacter{03B7}{\ensuremath{\eta}}
\DeclareUnicodeCharacter{0393}{\ensuremath{\Gamma}}
\DeclareUnicodeCharacter{0394}{\ensuremath{\Delta}}
\DeclareUnicodeCharacter{0398}{\ensuremath{\Theta}}
\DeclareUnicodeCharacter{039E}{\ensuremath{\Xi}}
\DeclareUnicodeCharacter{03A0}{\ensuremath{\Pi}}
\DeclareUnicodeCharacter{03C4}{\ensuremath{\tau}}
\DeclareUnicodeCharacter{03B3}{\ensuremath{\gamma}}
\DeclareUnicodeCharacter{03B4}{\ensuremath{\delta}}
\DeclareUnicodeCharacter{2205}{\ensuremath{\emptyset}}
\DeclareUnicodeCharacter{2261}{\ensuremath{\equiv}}
\DeclareUnicodeCharacter{2218}{\ensuremath{\circ}}
\DeclareUnicodeCharacter{1D539}{\ensuremath{\mathbb{B}}}
\DeclareUnicodeCharacter{1D553}{\ensuremath{\mathbb{b}}}
\DeclareUnicodeCharacter{220E}{\ensuremath{\blacksquare}}
\DeclareUnicodeCharacter{2208}{\ensuremath{\in}}
\DeclareUnicodeCharacter{22C6}{\ensuremath{\star}}
\DeclareUnicodeCharacter{231C}{\ensuremath{\ulcorner}}
\DeclareUnicodeCharacter{231D}{\ensuremath{\urcorner}}
\DeclareUnicodeCharacter{02B8}{\ensuremath{{}^y}}
\DeclareUnicodeCharacter{207F}{\ensuremath{{}^n}}
\DeclareUnicodeCharacter{1D3E}{\ensuremath{{}^P}}
\DeclareUnicodeCharacter{2219}{\ensuremath{\cdot}}

@@

\DeclareUnicodeCharacter{1D57}{\ensuremath{^t}}

@@

\renewcommand\Varid[1]{\mathord{\textsf{#1}}}

@@

\title{Can We Formalise Type Theory Intrinsically without Any Compromise?}
\subtitle{A Case Study in Cubical Agda}

\author{Liang-Ting Chen}
\orcid{0000-0002-3250-1331}
\affiliation{%
  \institution{Academia Sinica}
  \city{Taipei}
  \country{Taiwan}
}
\email{liangtingchen@as.edu.tw}

\author{Fredrik Nordvall Forsberg}
\orcid{0000-0001-6157-9288}
\affiliation{%
  \institution{University of Strathclyde}
  \city{Glasgow}
  \country{United Kingdom}
}
\email{fredrik.nordvall-forsberg@strath.ac.uk}

\author{Tzu-Chun Tsai}
\orcid{0009-0008-2855-5717}
\affiliation{%
  \institution{Academia Sinica}
  \city{Taipei}
  \country{Taiwan}
}
\email{gene0905@icloud.com}

@@

\keywords{Proof Assistants, Formalisation, Cubical Agda, Quotient Inductive-Inductive-Recursive Type}

@@

@@

\paragraph{Formalisation} Our formalisation can be found at \url{https://github.com/L-TChen/TTasQIIRT}, or archived with DOI \href{https://doi.org/10.5281/zenodo.17802827}{10.5281/zenodo.17802827}.

\section{Setting and Metatheory}

Our formalisation is carried out in \CA without the use of the \ensuremath{\Conid{Glue}} type and, in particular, the univalence principle, checked using the option \text{\ttfamily \char45{}\char45{}cubical\char61{}no\char45{}glue} available in the forthcoming \Agda 2.9.0. It also typechecks with \Agda version 2.8.0.
%
We explicitly set-truncate the types we define.
Therefore, the term \emph{set} is used interchangeably with type.
%
For the sake of simplicity we occasionally postulate uniqueness of identity proofs (UIP) locally, which is of course inconsistent with univalence but in principle compatible with cubical type theory.
%
We believe that the cubical type theory XTT~\cite{Sterling2022}, which enjoys definitional UIP without univalence, justifies this local assumption.
%

\CA implements cubical type theory, and one of the important concepts therein is the interval type \ensuremath{\Conid{I}} with two distinguished endpoints \ensuremath{\Varid{i0}} and \ensuremath{\Varid{i1}}.
%
Propositional equality \ensuremath{\Varid{x}\;\mathop\equiv\;\Varid{y}} in a type \ensuremath{\Conid{A}} is given by \emph{paths} in \ensuremath{\Conid{A}}, i.e., by a functions \ensuremath{\Conid{I}\;\Varid{→}\;\Conid{A}}.
More generally, dependent paths \ensuremath{\Varid{p}\;\mathbin{:}\;\Conid{PathP}\;\Conid{P}\;\Varid{a}\;\Varid{b}} are dependent functions \ensuremath{\Varid{p}\;\mathbin{:}\;(\Varid{i}\;\mathbin{:}\;\Conid{I})\;\Varid{→}\;\Conid{P}\;\Varid{i}} sending \ensuremath{\Varid{i0}} to \ensuremath{\Varid{a}\;\mathbin{:}\;\Conid{P}\;\Varid{i0}} and \ensuremath{\Varid{i1}} to \ensuremath{\Varid{b}\;\mathbin{:}\;\Conid{P}\;\Varid{i1}}.
Note that \ensuremath{\Conid{P}\;\mathbin{:}\;\Conid{I}\;\Varid{→}\;\Conid{Type}} itself is a path in the universe \ensuremath{\Conid{Type}}, hence a witness that \ensuremath{\Conid{P}\;\Varid{i0}\;\mathop\equiv\;\Conid{P}\;\Varid{i1}}, which the dependent path is \emph{over}, so paths are special cases of \ensuremath{\Conid{PathP}} with \ensuremath{\Conid{P}} being a constant path.
%
The constant path \ensuremath{\Varid{refl}}, defined as \ensuremath{\Varid{λ}\;\Varid{i}\;\Varid{→}\;\Varid{a}}, witnesses that equality is reflexive \ensuremath{\Varid{a}\;\mathop\equiv\;\Varid{a}}, and paths can be lifted to type families in the sense that there is a transport operation \ensuremath{\Varid{subst}\;\mathbin{:}\;(\Conid{P}\;\mathbin{:}\;\Conid{A}\;\Varid{→}\;\Conid{Type})\;\Varid{→}\;\Varid{x}\;\mathop\equiv\;\Varid{y}\;\Varid{→}\;\Conid{P}\;\Varid{x}\;\Varid{→}\;\Conid{P}\;\Varid{y}}.
%
See the literature on cubical type theory for details~\cite{Vezzosi2021}.
%

@@

\section{Type Theory as Quotient Inductive Types} \label{sec:tt}

@@

\subsection{Type Theory as the Initial CwF Model} \label{sec:tt:cwf}

@@

The four types of judgements in type theory are represented inductive-inductively

@@

When formulating the corresponding rule for the interaction between \ensuremath{\Varid{\char95 ∘\char95 }} and \ensuremath{\Varid{\char95 ,\char95 }}, we encounter a type mismatch that needs to be resolved by a transport (highlighted in \highlight{addition}{\text{green}}):
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{,∘}\;{}\<[5]%
\>[5]{}\mathbin{:}\;(\Varid{σ}\;\Varid{,}\;\Varid{t})\;\Varid{∘}\;\Varid{τ}\;\mathop\equiv\;(\Varid{σ}\;\Varid{∘}\;\Varid{τ}\;\Varid{,}\;\highlight{addition}{\Varid{subst}\;(\Conid{Tm}\;\Conid{Γ})\;[\circ]_{\text{T}}}\;(\Varid{t}\;[\kern-1.5pt\;\Varid{τ}\;\kern-1.5pt]_{\text{t}})){}\<[E]%
\ColumnHook
\end{hscode}\resethooks
The reason is that the type of \ensuremath{\Varid{t}\;[\kern-1.5pt\;\Varid{τ}\;\kern-1.5pt]_{\text{t}}} is \ensuremath{\Conid{A}\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{T}}\;[\kern-1.5pt\;\Varid{τ}\;\kern-1.5pt]_{\text{T}}} rather than the required \ensuremath{\Conid{A}\;[\kern-1.5pt\;\Varid{σ}\;\Varid{∘}\;\Varid{τ}\;\kern-1.5pt]_{\text{T}}}.
However, since \ensuremath{\Conid{Tm}} is an argument to \ensuremath{\Varid{subst}}, the use of transport violates a syntactic restriction of \Agda, namely its strict positivity check.
In theory, transports are allowed in QIITs~\cite{Kaposi2019}, but it is not clear to us how this syntactic restriction should be relaxed for higher inductive types supported by \CA to take into account other cubical primitives (such as \ensuremath{\Varid{hcomp}}).

@@

\subsection{The `Ford Transformation' and Index Elimination} \label{sec:tt:terms-without-indices}

To avoid transports in the definition itself, we note that the index \ensuremath{\Conid{A}} of \ensuremath{\Conid{Tm}\;\Conid{Γ}\;\Conid{A}} often needs an explicit proof for the typing constraint --- for example, that the term \ensuremath{\Varid{t}} in the substitution \ensuremath{(\Varid{σ}\;\Varid{,}\;\Varid{t})} has type \ensuremath{\Conid{A}\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{T}}} --- if this does not happen to hold strictly (i.e., up to judgemental equality), so enforcing this constraint in the index of \ensuremath{\Conid{Tm}} just shoots ourselves in the foot.
Hence, we apply the `Ford transformation'~\cite{McBride1999} (``You can have any index you want, as long as it is equal to the specified one'') to move the constraint on its index to its argument as an equality proof (highlighted below):
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{\char95 ,\char95 ∶[\char95 ]}\;\mathbin{:}\;(\Varid{σ}\;\mathbin{:}\;\Conid{Sub}\;\Conid{Γ}\;\Conid{Δ})\;(\Varid{t}\;\mathbin{:}\;\Conid{Tm}\;\Conid{Γ}\;\Conid{B})\;\highlight{addition}{(\Varid{pt}\;\mathbin{:}\;\Conid{B}\;\mathop\equiv\;\Conid{A}\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{T}})}\;{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{→}\;\Conid{Sub}\;\Conid{Γ}\;(\Conid{Δ}\;\Varid{,}\;\Conid{A}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks
The constructor \ensuremath{\Varid{,∘}} which had a transport in its type becomes
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{11}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{,∘}\;{}\<[5]%
\>[5]{}\mathbin{:}\;\Varid{...}\;(\Varid{σ}\;\Varid{,}\;\Varid{t}\;\Varid{∶[}\;\Varid{pt}\;\mskip1.5mu])\;\Varid{∘}\;\Varid{τ}\;{}\<[E]%
\\
\>[5]{}\hsindent{6}{}\<[11]%
\>[11]{}\mathop\equiv\;(\Varid{σ}\;\Varid{∘}\;\Varid{τ}\;\Varid{,}\;\Varid{t}\;[\kern-1.5pt\;\Varid{τ}\;\kern-1.5pt]_{\text{t}}\;\Varid{∶[}\;\highlight{addition}{\Varid{cong}\;\Varid{\char95 [}\;\Varid{τ}\;\kern-1.5pt]_{\text{T}}\;\Varid{pt}\;\Varid{∙}\;[\circ]_{\text{T}}}\;\mskip1.5mu]){}\<[E]%
\ColumnHook
\end{hscode}\resethooks
where \ensuremath{\Varid{\char95 ∙\char95 }} is the transitivity of equality.
Although transport is not needed this time, the use of \ensuremath{\Varid{cong}} and \ensuremath{\Varid{\char95 ∙\char95 }}
still prevent the definition from being seen as strictly positive.
Similar to the Ford transformation, this problem can be overcome by asking for another equality proof, highlighted below, as an argument:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{4}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{,∘}\;\mathbin{:}\;\Varid{...}\;\highlight{addition}{(\Varid{qt}\;\mathbin{:}\;\Conid{B}\;[\kern-1.5pt\;\Varid{τ}\;\kern-1.5pt]_{\text{T}}\;\mathop\equiv\;\Conid{A}\;[\kern-1.5pt\;\Varid{σ}\;\Varid{∘}\;\Varid{τ}\;\kern-1.5pt]_{\text{T}})}\;{}\<[E]%
\\
\>[B]{}\hsindent{4}{}\<[4]%
\>[4]{}\Varid{→}\;(\Varid{σ}\;\Varid{,}\;\Varid{t}\;\Varid{∶[}\;\Varid{pt}\;\mskip1.5mu])\;\Varid{∘}\;\Varid{τ}\;\mathop\equiv\;(\Varid{σ}\;\Varid{∘}\;\Varid{τ})\;\Varid{,}\;\Varid{t}\;[\kern-1.5pt\;\Varid{τ}\;\kern-1.5pt]_{\text{t}}\;\Varid{∶[}\;\highlight{addition}{\Varid{qt}}\;\mskip1.5mu]{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
As \ensuremath{\Conid{Sub}} is a set, the additional argument is essentially unique, so this updated constructor does not require any additional information but only defers the proof obligation.
